---
title: "Intro to time series wrangling & viz"
author: "Roshni Katrak-Adefowora"
date: "1/31/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(lubridate)
library(tsibble)
library(feasts)
library(slider)
```

### Read in the data
```{r}
toolik <- read_csv(here("data", "toolikweather.csv"))
```

### Convert this into a tsibble (time series dataframe)
```{r}
toolik_ts <- toolik %>% 
  mutate(date = mdy(date)) %>% 
  as_tsibble(key = NULL, index = date) #no key because all the sites are the same, so no obvious grouping
```

```{r}
ggplot(data=toolik_ts, aes(x=date, y=mean_airtemp))+
  geom_line()

#see strong seasonality, don't really see trend or cyclicality; also no big outliers or gaps in the data
```

### Use `index_by` function (time series equivalent of `group_by()`) to aggregate time series info by intervals we specify
```{r}
toolik_month <- toolik_ts %>% 
  index_by(yr_mo = ~yearmonth(.)) %>%  #aggregate by year and month across entire time series
  summarize(monthly_mean_temp = mean(mean_airtemp, na.rm = TRUE))

ggplot(data=toolik_month, aes(x=yr_mo, y=monthly_mean_temp))+
  geom_line()+
  facet_wrap(~month(yr_mo, label = TRUE)) #use month function from lubridate to pull month from yr_mo column
```
Aggregate by week:
```{r}
toolik_weekly <- toolik_ts %>% 
  index_by(weekly = ~yearweek(.)) %>% 
  summarize(weekly_airtemp = mean(mean_airtemp, na.rm = TRUE),
            min_airtemp = min(mean_airtemp))

head(toolik_weekly) #see first 6 rows
```

### Filtering time series

Use `filter_index()` to specify ranges of time series to keep or exclude.
```{r}
toolik_ts %>% 
  filter_index("2000-06" ~ "2001-10")
```

```{r}
toolik_ts %>% 
  filter_index("2006-04-10" ~ "2006-05-15")
```

### Seasonlots
```{r}
toolik_ts %>% 
  filter(year(date) > 2014) %>% #could also do filter_index("2015" ~ .)
  gg_season(y = mean_airtemp)

toolik_month %>% 
  gg_season(y = monthly_mean_temp)+
  theme_light()
```

### Seasonal subseries plot
```{r}
toolik_month %>% 
  gg_subseries(monthly_mean_temp)
```

### Find moving average

Using the slider package in R
```{r}
set.seed(2021)
test <- rnorm(n=100, mean=40, sd=10)

#look at the values for the averages
slide(test, ~.x, .before = 2, .after = 2) #across all values of vector x, extends to 2 values before and 2 values after
```

```{r}
roll_toolik_15 <- toolik_ts %>% 
  mutate(ma_15d = as.numeric(slide(toolik_ts$mean_airtemp, #stored as numeric instead of list
                                   mean,
                                   .before = 7,
                                   .after = 7)))

ggplot()+
  geom_line(data=toolik_ts, aes(x=date, y=mean_airtemp), size=0.2, color="gray")+
  geom_line(data=roll_toolik_15, aes(x=date, y=ma_15d))+
  theme_minimal()
```

### Create the ACF (visual representation of how observations are correlated with prior observations on the same variable, over specified lags)
```{r}
#for days - pretty correlated
toolik_ts %>% 
  ACF(mean_airtemp) %>% 
  autoplot()

#for months - see observations 12 months apart are strongly positively correlated. 6 or 18 months apart are negatively correlated.
toolik_month %>% 
  ACF(monthly_mean_temp) %>% 
  autoplot()
```

### Decomposition by STL
```{r}
toolik_dec <- toolik_month %>% 
  model(STL(monthly_mean_temp ~ season(window = Inf)))

components(toolik_dec) %>% 
  autoplot()
```

